---
title: "Spatio-Temporal Kriging"
author: "Kate Pyper"
output:
  xaringan::moon_reader:
    css: ["xaringan-themer.css"]
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r, include = FALSE}
library(tidyverse)
library(gstat)
library(sf)
library(spacetime)

dat <- read_csv("/conf/quality_indicators/Climate/data/ephss_data/PHS request May 2024/295 mlem_max temperature.csv", skip = 2) %>% 
  mutate(Time = dmy(str_sub(Time, 1, 10))) %>% 
  filter(year(Time) == 2022)
```

## This session will look at

We'll start by looking at some of the theory behind kriging:

- Variograms 
- Kriging algorithm
- Some additional things to consider while kriging

Then we'll move on to looking at the code itself. The structure will be outlined in these slides, but we will move to R to discuss the finer details.

---
class: inverse, center, middle

# Kriging

---

## Spatial and temporal modelling

When looking at a time series, we are generally pretty comfortable with the idea that things change gradually over time 

*Things which are closer together in time are more similar than things which are further apart in time.*

This gives way to the concept of temporal autocorrelation.

A similar phenomenon can be observed across space, *where things which are close together in space are more similar than things which are further apart*.

For example, the weather you're experiencing is likely to be more similar to the weather just down the street than the weather on the other side of the country. 

This is spatial autocorrelation.

Kriging leverages autocorrelation in order to make predictions at new locations, but in order to do that we need to know something about what the correlation looks like - for that we use a *variogram*.

---

## Spatial variograms

Focusing initially on spatial analysis only, a sample variogram splits the range of distances up into ranges, and within each range calculates the average squared difference between pairs of observations which are separated by a distance within that range. 

```{r, echo = FALSE, fig.width = 7, fig.height = 4, fig.align='center'}
dat2 <- filter(dat, Time == min(Time))
sample_variogram <- variogram(Value ~ 1, data = dat2, locations = ~Longitude + Latitude)
plot(sample_variogram)
```

Variograms will typically have this same shape, where there is an initial increase in the value of the variogram followed by a levelling off. Note that we are more interested in the left hand end of the variogram than the right hand end, as these points are generally based on a very small number of observations.

---

## Spatial variogram models

The sample variogram allows us to see the approximate correlation structure within our data, but isn't particularly useful beyond that. In order to allow us to evaluate the correlation between two observations at any separating distance we need to use a model. 

There are multiple types of variogram model available, but the main ones are exponential, spherical and (for more flexibility). These are generally of a similar (though not identical) shape, and all make use of the following parameters:

- Nugget - discontinuity at the origin

- Sill - maximum value of the variogram

- Partial Sill - Sill $-$ Nugget

- Range - The separating distance at which the Sill is reached

We fit our variogram model to the sample variogram via an optimisation algorithm, so we want to aim to specify these parameters as closely as possible to make sure the model reaches convergence.

---

## Spatio-temporal Variograms

Similar to a spatial variogram, but with the added time dimension can be viewed either as a heat map or as separate traces for each time point

```{r include = FALSE}

spat <- st_as_sf(data.frame(long = unique(dat$Longitude), 
                            lat = unique(dat$Latitude)), 
                 coords = c("long", "lat"), crs = 4326) %>% 
  as_Spatial()

# unique time points 
time <- unique(dat$Time)

# Index grid to say which sites are observed at which times - indexes spat (col 1) and time (col 2)
index <- select(dat, `Site identifier`, Time) %>% 
  mutate(site = map_int(`Site identifier`, \(x) which(unique(`Site identifier`) == x)), 
         ob_date = map_int(Time, \(x) which(unique(Time) == x)),
         .keep = "unused")

# Extract relevant pieces of data - remove coordinates and time as these are present in the object
dat1  <- select(dat, -c(Latitude, Longitude))

# Set up spacetime object for locations 
# - STSDF because not every location has an observation at every time point
dat_st <- STSDF(spat, time, dat1, as.matrix(index))

sample_variogram <- variogramST(Value ~ 1, data = dat_st)

```

```{r, echo = FALSE, fig.width = 10, fig.height = 4, fig.align = 'center'}
gridExtra::grid.arrange(plot(sample_variogram), plot(sample_variogram, map = FALSE), ncol = 2)
```


Both plots show the same variogram in different ways. If looking at something similar to the left, you would expect to see darker colours on the bottom and left hand side, and colours gradually turning more warm the further up and to the right of the plot you go. In the plot on the right, wach individual line should look like a regular spatial variogram. Variograms will gradually start to group together at higher lags. 

---

## Spatio-temporal variogram models

For prediction, we will need to use a Spatio-temporal variogram model. These models will all require a spatial component and a temporal component.  More complex spatio-temporal variogram models also require a joint component.

Each of the model components has the same form as a spatial variogram model. As in the spatial only model, we should aim to have a good guess at the parameters to start off. If using a joint component make the nugget for the spatial and temporal components 0.

### Temporal component

- Partial Sill - at what value on the y axis do the separate variograms start to clump together

- Range - at what lag do the separate variograms start to clump together

### Spatial component

- Partial Sill - height of the individual variograms

- Range - range of the individual variograms

### Joint component

- Partial Sill - overall distance between nugget and the partial sill of the temporal model

- Range - distance times time lag at which the surface is levelling off

---

## Variograms and covariance

The variogram describes how different pairs of observations are at various separating distances. The variogram is directly related to the covariance:

- The Sill is the variance

- $\text{Sill} - \gamma(h)$ is the covariance between two observations at locations which are $h$ units apart

The kriging process uses the variogram to obtain covariances, which are then used to weight observations when making predictions at new locations. 

---

## Kriging

There are three types of kriging:

- Simple Kriging - a fixed constant mean is supplied to the process

- Ordinary Kriging - a constant mean is estimated and used in the process

- Universal Kriging - There are covariates involved in the prediction process

For this project we don't have covariates, so we're going to focus on Ordinary Kriging.

---

## Ordinary Kriging

Kriging works by taking a weighted average of observations, where the weights are based on the covariances present within the observations. Specifically the weights are:

$$ \boldsymbol{\lambda} = \boldsymbol{A}^{-1}\boldsymbol{b} $$
Where $\boldsymbol{A}$ is the covariance matrix of the observations involved in the prediction, and $\boldsymbol{b}$ is the covariance between the location that we're trying to predict at and the observations being used to make the prediction. 
---

## Things to consider

A couple of things that we have had to think about as part of this process:

### Local Kriging

Since we wanted to retain as much of the variability in weather across the country as possible, doing a global kriging meant that we were over-smoothing, so we decided to implement a local kriging procedure. When running for a new weather variable, make sure the level of "local-ness" is giving sufficient variability relative to the observations. More local means more variability.

### Transformations

Kriging works best when we have normally distributed data that covers the entire real line. This will not be true for all weather variables. When considering transformations, think about trying to make the observations as symmetric as possible - it may help to think about some common transformations (e.g. relative humidity is a percentage, so we were able to use a logistic transformation)

---

class: center, middle, inverse2

# The Code

---

# Script to generate predictions

Each weather indicator has a separate generation script. The script is split into two parts:

- Part 1 reads in the data, sources the functions required and runs the full procedure

- Part 2 (`Check Data`) exists to allow you to take a look at the data before you run the full procedure. Use this to check whether transformations are needed, and to get a feel for what your variogram parameters should be.

---

# Functions

There are 4 functions that are required to run the kriging procedure, and they are nested within each other.

- `krige_st()` - takes a quarter and outputs a csv file containing the results of the kriging. This calls the function `preds()`.

- `preds()` - takes an intermediate zone, a fitted variogram, the spatio-temporal version of the data, and the data itself and returns predictions for that intermediate zone. This calls the function `preds_single()`.

- `preds_single()` - takes a data frame, an intermediate zone, a time point, and a fitted variogram model and returns a prediction for an IZ at that time point. This calls the function `vgm_cov()`.

- `vgm_cov()` - takes a site, a data frame that contains distances, and a fitted variogram model and calculates estimates of the covariance from the fitted variogram model.


